{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-29 ms.tex
Reworking of metapopulation section

Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
enables multiple complex non-linear demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}).

Second, {\plant} allows for the effects of competition and succession on
productivity and biomass accumulation within a patch to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with patch age. Such changes might come about via a
combination of physiological and structural changes
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors. Moreover, {\plant} allows for the
contributions of different species to be mapped (Fig. \ref{fig:patch}c).

@@

Most vegetation is subject to a disturbance regime, such that patches of
the landscape are cleared at some rate, by fire, cyclone, landslide, or
disease outbreak
\citep{White-1979, Bormann-1979, Chambers-2013}.
The vegetation thus comprises a collection of patches differing in time
since last disturbance and linked via seed dispersal (Fig.
\ref{fig:schematic}b). Such a system is often refereed to as a
structured metapopulation \citep{Gyllenberg-2001}.

In {\plant} we assume an infinite number of patches all experiencing
the same disturbance regime; we also assume that the disturbances remove
all established vegetation.  With this approach we can scale up from
simulating a single patch to simulate an entire metapopulation.
The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second \textsc{pde}
\citep{ Vonfoerster-1959} (see Appendix
\ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation.

@@

As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
feature of the size-structured metapopulation concept implemented here
is that it balances equilibrium and non-equilibrium approaches to
modelling ecological dynamics. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

\subsection{Applications}
% TODO[RGF]:  But we need
% something that indicates how this can be used.  I suggest a complete
% overhaul of this paragraph as that gets closest to a way of
% using the model.

Beyond its conceptual appeal, this idea of a dynamic equilibrium
allows us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits,
competitive interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of
patches. Unlike some vegetation models, there is no need to set the
density of seeds or seedlings, these arise as outputs from the
model. Characteristics of the vegetation such as total leaf area,
biomass, productivity and allocation to different plant parts also
arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
distributions of abundance, leaf area and growth rate with respect to
size emerge at both patch scale and scale of metapopulation
(Fig. \ref{fig:emergent}). Note in particular how a single
physiological model produces a variety of height-density
relationships, dependent on patch age, as well as an emergent average
across all patches (Fig. \ref{fig:emergent}, see more detail in
Appendix \ref{sec:emergent}). This emergent relationship might then be
compared to the observed distributions in large forest plots
\citep{Muller-2006}.

\section{Invasion fitness}

With the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. A new type can invade when
\(R > 1\).

@@

>>>>

Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
enables multiple complex non-linear demographic phenomena to be
investigated. First, {\plant} can simulate patterns of patch development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
 species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}). Second, {\plant} allows for the effects of competition and succession on
productivity and biomass accumulation within a patch to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with patch age. Such changes might come about via a
combination of physiological and structural changes
\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors. Moreover, {\plant} allows for the
contributions of different species to be mapped (Fig. \ref{fig:patch}c).

@@

Most vegetation is subject to a disturbance regime, such that patches of
the landscape are cleared at some rate, by fire, cyclone, landslide, or
disease outbreak
\citep{White-1979, Bormann-1979, Coomes-2007, Chambers-2013}.
The vegetation thus comprises a collection of patches differing in time
since last disturbance and linked via seed dispersal (Fig.
\ref{fig:schematic}b). Such a system is often refereed to as a
structured metapopulation \citep{Gyllenberg-2001}.

In {\plant} we assume an infinite number of patches, all experiencing
the same disturbance regime and linked via seed dispersal (Fig.
\ref{fig:schematic}b). We also assume that the disturbances remove
all established vegetation.  With this approach we can scale up from
simulating a single patch to simulate an entire metapopulation.
The frequency-density
\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second \textsc{pde}
(see Appendix \ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches aged \(a\) in the metapopulation.

@@

A feature of the size-structured metapopulation concept implemented here
is that it balances equilibrium and non-equilibrium approaches to
modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001, 
Falster-2011}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

@@

Beyond its conceptual appeal, the idea of a dynamic equilibrium captured by
the trait-, size- and patch structured metapopulation model potentially
resolves several challenges faced by unstructured models. First, by scaling up
from patch to metapopulation, we close the demographic feedback loop to create
a self-regulating system. Unlike some vegetation models, there is no need to
set the density of seeds or seedlings, these arise as outputs from the model,
based on the combined effects of the model's physiological rules, species
traits, competitive interactions and the prevailing disturbance regime.

Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches are more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c). Moreover, predictions are for a frequency distribution 
of states across the landscape. Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} may
can be modelled as  emergent phenomena, where variety of patch-level
relationships combine to  produce  an emergent average across all patches
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size-and patch- structured models may be more suitable for comparing with 
satellite and survey data, than unstructured models representing the world as 
homogeneous green slime \citep{Moorcroft-2001,Moorcroft-2001}.

Finally, with the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. A new type can invade when
\(R > 1\).
